draft

### 2025-05-20 draft.tex
docs: add imports

% Created 2025-05-19 Mon 21:10
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

By separating the numerical from the user-facing and orchestration components we improved reusability, testability and extensibility. We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to perform data analyses, while still retaining the standard features already present in numerous datasets.

@@

The validity of cp\_measure is shown on Figure \ref{fig:cp_vs_cpmeasure}. We show examples of the features on the panel on the left. Mostly straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which may indicate that some edge-cases are treated differently. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements resulted in a linear relation, regardless of whether the masks were for nuclei or cytosols. This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
\subsection{Results}
\label{sec:orge5b5c6b}
We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

As a demonstration of its ease of use and the value of interpretable features, we used cp\_measure in a standard classification workflow. We processed 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}. We preprocessed the data following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired. With this we identified which features distinguish cells on the later samples and distinguish subpopulations. Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells [CITE Shapley].

@@

To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy (CITE). Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy. Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts. Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.

To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data. This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization. We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/). Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA'). Morphological features were extracted from these 5-channel images for both datasets. Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and Pycytominer (CITE) respectively. We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig. XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data. Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type. Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included. Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.

>>>>

% Created 2025-05-20 Tue 10:41
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

By isolating the numerical from the graphical and orchestration components we improved reusability, testability and extensibility. We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already present in numerous datasets.

@@

The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols. This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
\subsection{Results}
\label{sec:orge5b5c6b}
We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

As a demonstration of its ease of use and the value of interpretable features, we used cp\_measure in a standard classification workflow. We processed 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}. We preprocessed the data following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired. With this we identified which features distinguish cells on the later samples and distinguish subpopulations. Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.

@@

To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy \citep{pallaSquidpyScalableFramework2022}. Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy. Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts. Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.

To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data. This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization. We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform \citep{CosMxSMIMouse2025}. Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA'). Morphological features were extracted from these 5-channel images for both datasets. Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy \citep{wolfSCANPYLargescaleSinglecell2018} and Pycytominer \citep{serranoReproducibleImagebasedProfiling2025} respectively. We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig. XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data. Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type. Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included. Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.
